---
title: "Advanced Matching Techniques"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Matching Techniques}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The basic `dmd_price_lookup()` function supports three match methods: `"partial"`, `"exact"`, and `"fuzzy"`. This vignette covers advanced techniques for difficult matching scenarios.

```{r setup}
library(dmdprices)
library(dplyr)
library(stringr)
```

## The three built-in methods

### 1. Partial matching (default)

Matches any medicine name containing all your search terms, case-insensitive.

```{r}
# Very forgiving
dmd_price_lookup("metformin", method = "partial")
```

**Use when:**
- You don't know the exact name
- You want all related products
- Searching broad categories

### 2. Exact matching

Requires the full VMP (Virtual Medicinal Product) name.

```{r}
# Strict—must match the canonical dm+d name exactly
dmd_price_lookup("Metformin 500mg tablets", method = "exact")
```

**Use when:**
- You have clean, standardized data
- You want to avoid false matches
- Building reproducible workflows

### 3. Fuzzy matching

Uses Levenshtein distance to tolerate typos and minor variations.

```{r}
# Tolerates ~2-3 character differences
dmd_price_lookup("metformin 500 mg tabets", method = "fuzzy", max_dist = 3)
```

**Parameters:**
- `max_dist`: Maximum character distance (default 2)
  - Lower = stricter (fewer false matches)
  - Higher = more permissive (catches typos)

## Advanced scenario 1: Matching branded to generic names

You have a list of branded products but need generic (VMP) prices.

```{r}
# Branded product name from prescriptions
branded_medicines <- c(
  "Glucophage (Metformin 500mg tablets)",
  "Lisinopril 10mg tablets (various)",
  "Atorvastatin 20mg tablets (Actavis)"
)

# Extract the core medicine (parentheses are noise)
core_names <- str_extract(branded_medicines, "^[^(]+")  # Text before (
core_names <- str_trim(core_names)

# Look up the generic VMP
results <- lapply(core_names, function(med) {
  dmd_price_lookup(med, method = "fuzzy", max_dist = 3)
})

# Extract the canonical VMP names
vmp_names <- sapply(results, function(r) {
  if (nrow(r) > 0) r$medicine[[1]] else NA_character_
})

vmp_names
```

## Advanced scenario 2: Standardizing supplier codes to medicines

You have internal supplier codes and need to match to dm+d.

```{r}
# Internal mapping table
supplier_codes <- data.frame(
  supplier_code = c("SUPP001", "SUPP002", "SUPP003"),
  our_name = c(
    "paracetamol 500mg",
    "ibuprofen 200mg",
    "aspirin 75mg"
  )
)

# Match using fuzzy matching
supplier_codes <- supplier_codes |>
  mutate(
    dmd_match = sapply(our_name, function(med) {
      result <- dmd_price_lookup(med, method = "fuzzy", max_dist = 3)
      if (nrow(result) > 0) result$medicine[[1]] else NA_character_
    }),
    dmd_price_pence = sapply(our_name, function(med) {
      result <- dmd_price_lookup(med, method = "fuzzy", max_dist = 3)
      if (nrow(result) > 0) result$basic_price[[1]] else NA_real_
    })
  )

supplier_codes
```

## Advanced scenario 3: Batch processing with quality checks

Processing many medicines and tracking which matched poorly.

```{r}
# Large medicine list (simulated)
medicine_list <- c(
  "Metformin 500mg tablets",
  "Lisinopril 10mg tablets",
  "Amlodipine 5mg tablets",
  "Fluoxetine 20mg capsules",
  "Omeprazole 20mg capsules",
  "xyz_fake_medicine"  # Will fail to match
)

# Batch lookup with error handling
batch_results <- data.frame(
  input_name = medicine_list,
  matched_vmp = NA_character_,
  price_pence = NA_real_,
  match_confidence = NA_character_,
  matched = NA
)

for (i in seq_along(medicine_list)) {
  result <- dmd_price_lookup(
    medicine_list[i],
    method = "fuzzy",
    max_dist = 3
  )
  
  if (nrow(result) > 0) {
    batch_results$matched_vmp[i] <- result$medicine[[1]]
    batch_results$price_pence[i] <- result$basic_price[[1]]
    batch_results$match_confidence[i] <- "High"  # fuzzy matched
    batch_results$matched[i] <- TRUE
  } else {
    batch_results$matched[i] <- FALSE
  }
}

batch_results
```

**Quality report:**

```{r}
batch_results |>
  summarise(
    total_medicines = n(),
    matched = sum(matched, na.rm = TRUE),
    unmatched = sum(!matched, na.rm = TRUE),
    match_rate = paste0(round(100 * mean(matched), 1), "%")
  )
```

## Advanced scenario 4: Handling strength/form ambiguity

A medicine exists in multiple strengths; you need the right one.

```{r}
# Partial search returns all strengths
all_atorvastatin <- dmd_price_lookup("atorvastatin", method = "partial")

all_atorvastatin |>
  select(medicine, pack_size, unit, basic_price) |>
  slice(1:6)  # Show first 6
```

**Filter to the strength you want:**

```{r}
# You specifically want 20mg tablets
target <- all_atorvastatin |>
  filter(
    str_detect(medicine, "20mg"),
    str_detect(medicine, "tablets")
  )

target |>
  select(medicine, basic_price)
```

## Advanced scenario 5: Tracking price changes

Look up the same medicine over time with version control.

```{r}
# Suppose you've recorded metformin prices monthly
# (This is simulated; in reality you'd load old dm+d versions)
price_history <- data.frame(
  month = c(
    "2023-01", "2023-02", "2023-03",
    "2024-01", "2024-02", "2024-03"
  ),
  price_pence = c(
    145, 147, 148,
    152, 155, 158
  )
)

# Add pence-to-pound conversion
price_history <- price_history |>
  mutate(price_pounds = price_pence / 100)

price_history
```

**Calculate month-on-month change:**

```{r}
price_history <- price_history |>
  mutate(
    previous_price = lag(price_pence),
    change_pence = price_pence - previous_price,
    pct_change = round(100 * (change_pence / previous_price), 2),
    .after = price_pence
  )

price_history |>
  select(month, price_pence, change_pence, pct_change)
```

## Advanced scenario 6: Fuzzy matching with manual review

For critical systems, allow fuzzy matches but flag for review.

```{r}
# Function: fuzzy match with confidence scoring
fuzzy_match_scored <- function(medicine_name, db = dmd_master, max_dist = 2) {
  result <- dmd_price_lookup(medicine_name, method = "fuzzy", max_dist = max_dist, db = db)
  
  if (nrow(result) == 0) {
    return(data.frame(
      input = medicine_name,
      matched = FALSE,
      vmp_name = NA_character_,
      price = NA_real_,
      requires_review = TRUE,
      reason = "No match found"
    ))
  }
  
  # Flag if multiple matches or high Levenshtein distance
  requires_review <- nrow(result) > 1 || max_dist > 2
  
  data.frame(
    input = medicine_name,
    matched = TRUE,
    vmp_name = result$medicine[[1]],
    price = result$basic_price[[1]],
    requires_review = requires_review,
    reason = if (nrow(result) > 1) "Multiple matches" else "Fuzzy match"
  )
}

# Test
test_medicines <- c(
  "Metformin 500mg tablets",
  "paracetmol 500mg",  # Typo
  "xyz_fake"
)

matches <- lapply(test_medicines, fuzzy_match_scored) |>
  bind_rows()

matches
```

**Filter for review:**

```{r}
matches |>
  filter(requires_review | !matched) |>
  select(input, matched, reason)
```

## Advanced scenario 7: Partial matching with results filtering

Use partial search but filter results intelligently.

```{r}
# You want all ibuprofen products, but not creams/gels
all_ibuprofen <- dmd_price_lookup("ibuprofen", method = "partial")

# Filter to oral forms only
oral_ibuprofen <- all_ibuprofen |>
  filter(
    unit %in% c("tablet", "capsule", "liquid", "suspension"),
    !str_detect(str_to_lower(medicine), "cream|gel|topical|ointment")
  )

oral_ibuprofen |>
  select(medicine, pack_size, unit, basic_price) |>
  arrange(medicine)
```

## Best practices for difficult matches

### 1. Always define your matching logic upfront

```r
matching_rules <- list(
  method = "fuzzy",           # Use fuzzy for typo tolerance
  max_dist = 2,               # Strict but reasonable
  exclude_patterns = c(
    "cream", "gel",           # Exclude topicals
    "injection",              # Only interested in oral
    "solution"
  ),
  include_only = c(
    "tablets", "capsules",
    "suspension", "liquid"
  )
)
```

### 2. Document match quality in your results

```{r}
# Add a "match_quality" column
quality_results <- batch_results |>
  mutate(
    match_quality = case_when(
      !matched ~ "unmatched",
      is.na(price_pence) ~ "no_price",
      TRUE ~ "good"
    ),
    .after = matched
  )

quality_results
```

### 3. Log unmatched items for investigation

```r
unmatched_log <- batch_results |>
  filter(!matched) |>
  mutate(
    investigated = FALSE,
    notes = NA_character_
  ) |>
  select(input_name, investigated, notes)

# Save for manual review
write.csv(unmatched_log, "unmatched_medicines_review.csv", row.names = FALSE)
```

### 4. Test on a small sample first

```r
# Before processing thousands, validate on 20-30
test_sample <- head(medicine_list, 10)
test_results <- lapply(test_sample, dmd_price_lookup, method = "fuzzy")
# Review results manually before proceeding
```

## Comparing match methods

```{r}
test_name <- "metformin 500 mg"

# Exact
exact <- dmd_price_lookup(test_name, method = "exact")
cat("Exact:", nrow(exact), "results\n")

# Partial
partial <- dmd_price_lookup(test_name, method = "partial")
cat("Partial:", nrow(partial), "results\n")

# Fuzzy
fuzzy <- dmd_price_lookup(test_name, method = "fuzzy", max_dist = 2)
cat("Fuzzy:", nrow(fuzzy), "results\n")
```

## Performance tips

For large batches (1000+ medicines):

```r
# Use vectorized operations where possible
medicines <- c("medicine1", "medicine2", ..., "medicine1000")

# Faster: use sapply with built-in parallelization
system.time({
  results <- sapply(medicines, dmd_price_lookup, method = "exact")
})

# For very large batches, consider:
# 1. Splitting into chunks of 100-200
# 2. Using parallel::lapply() with multiple cores
# 3. Pre-filtering your list to remove obvious non-matches
```

## Further reading

- Basic usage: `vignette("dmdprices")`
- Price reconciliation: `vignette("drug_tariff_matching")`
- Troubleshooting: `vignette("troubleshooting")`
