---
title: "Cost Analysis Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cost Analysis Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette shows how to combine `dmd_price_lookup()` and `inflate_nhscii()` to track medicine costs over time and across different financial years.

```{r setup}
library(dmdprices)
library(dplyr)
```

## Scenario: Tracking a medicine's cost over time

Suppose you're evaluating the cost of **metformin 500mg tablets** across multiple years in your pharmacy budget.

### Step 1: Look up the current price

```{r}
metformin <- dmd_price_lookup("metformin 500mg tablets", method = "exact")
metformin
```

Extract the price:

```{r}
current_price_pence <- metformin$basic_price[[1]]
current_price_pounds <- current_price_pence / 100

cat("Current price (Drug Tariff):", current_price_pounds, "pounds\n")
```

### Step 2: Estimate historical costs using NHS CII

You want to know what this medicine would have cost in previous years (in comparable prices):

```{r}
# Current year is 2023/24
# Deflate to earlier years
years <- c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24")

historical_costs <- data.frame(
  year = years,
  deflated_price_pence = NA_real_
)

for (i in seq_along(years)) {
  deflation_factor <- nhscii(
    from_year = "2023/24",
    to_year = years[i],
    index = "prices"
  )
  historical_costs$deflated_price_pence[i] <- current_price_pence * deflation_factor
}

historical_costs <- historical_costs |>
  mutate(
    deflated_price_pounds = deflated_price_pence / 100,
    .before = deflated_price_pence
  )

historical_costs
```

**Interpretation:** This shows what metformin would have cost in each year if prices had moved with the NHS CII "prices" index.

### Step 3: Project future costs

Now inflate across the available data and estimate future years. Note that 2024/25 and 2025/26 rates will be published in the PSSRU 2025 manual; for now we use the 2023/24 rate as a baseline:

```{r}
# Available historical years
historical_years <- c("2020/21", "2021/22", "2022/23", "2023/24")

historical_projection <- data.frame(
  year = historical_years,
  projected_price_pounds = NA_real_
)

for (i in seq_along(historical_projection$year)) {
  inflation_factor <- nhscii(
    from_year = "2023/24",
    to_year = historical_projection$year[i],
    index = "prices"
  )
  historical_projection$projected_price_pounds[i] <- current_price_pounds * inflation_factor
}

historical_projection

# For future projections, you would apply estimated inflation factors
# once 2024/25 and 2025/26 NHS CII rates are published
inflation_estimate_2024_25 <- 1.04  # Estimated at 4% based on recent trends
inflation_estimate_2025_26 <- 1.03  # Estimated at 3%

future_projection <- data.frame(
  year = c("2024/25", "2025/26"),
  projected_price_pounds = c(
    current_price_pounds * inflation_estimate_2024_25,
    current_price_pounds * inflation_estimate_2024_25 * inflation_estimate_2025_26
  )
)

future_projection
```

## Scenario: Budget impact analysis

You need to estimate the budget impact of a medicine switch across multiple medicines.

### Setup: Your current medicines

```{r}
# Current medicines in your formulary (fictitious costs)
current_formulary <- data.frame(
  medicine = c(
    "Metformin 500mg tablets",
    "Lisinopril 10mg tablets",
    "Atorvastatin 20mg tablets"
  ),
  current_monthly_packs = c(500, 300, 250)
)

# Look up each medicine
current_formulary$price_pence <- sapply(
  current_formulary$medicine,
  function(med) {
    result <- dmd_price_lookup(med, method = "fuzzy", max_dist = 2)
    if (nrow(result) > 0) {
      result$basic_price[[1]]
    } else {
      NA_real_
    }
  }
)

current_formulary <- current_formulary |>
  mutate(price_pounds = price_pence / 100)

current_formulary
```

### Calculate monthly and annual costs

```{r}
current_formulary <- current_formulary |>
  mutate(
    monthly_cost = current_monthly_packs * price_pounds,
    annual_cost_2023_24 = monthly_cost * 12,
    .after = price_pounds
  )

current_formulary
```

### Project costs to 2024/25

```{r}
inflation_factor_2024_25 <- nhscii(
  from_year = "2023/24",
  to_year = "2023/24",  # We don't have 2024/25 yet; for demo use current year
  index = "pay_and_prices"
)

# For projection purposes (assuming modest inflation)
inflation_estimate_2024_25 <- 1.04  # 4% placeholder

current_formulary <- current_formulary |>
  mutate(
    annual_cost_2024_25 = annual_cost_2023_24 * inflation_estimate_2024_25,
    cost_increase = annual_cost_2024_25 - annual_cost_2023_24
  )

current_formulary |>
  select(medicine, annual_cost_2023_24, annual_cost_2024_25, cost_increase)
```

**Summary:**

```{r}
current_formulary |>
  summarise(
    total_2023_24 = sum(annual_cost_2023_24, na.rm = TRUE),
    total_2024_25 = sum(annual_cost_2024_25, na.rm = TRUE),
    total_increase = sum(cost_increase, na.rm = TRUE),
    pct_increase = (total_increase / total_2023_24) * 100
  )
```

## Scenario: Cost-effectiveness analysis with historical comparison

Evaluate how a medicine's cost-effectiveness ratio changes with inflation adjustment.

```{r}
# Simulated QALY data and costs
cea_data <- data.frame(
  medicine = c("Medicine A", "Medicine B"),
  cost_2020_21_pounds = c(5000, 7500),
  qalys = c(0.50, 0.75)
)

# Inflate costs to 2023/24 (current year)
cea_data <- cea_data |>
  mutate(
    inflation_factor = nhscii(
      from_year = "2020/21",
      to_year = "2023/24",
      index = "pay_and_prices"
    ),
    cost_2023_24_pounds = cost_2020_21_pounds * inflation_factor,
    cost_per_qaly_2020_21 = cost_2020_21_pounds / qalys,
    cost_per_qaly_2023_24 = cost_2023_24_pounds / qalys
  )

cea_data |>
  select(
    medicine, 
    cost_per_qaly_2020_21, 
    cost_per_qaly_2023_24
  )
```

## Scenario: Comparing costs across NHS trusts

Different trusts may purchase medicines at different times. Normalize to a common year:

```{r}
# Data from multiple trusts (purchased in different years)
trust_data <- data.frame(
  trust = c("Trust A", "Trust B", "Trust C"),
  medicine = "Paracetamol 500mg tablets",
  purchase_year = c("2021/22", "2022/23", "2023/24"),
  cost_paid_pounds = c(150, 180, 220),
  quantity = c(1000, 1200, 1500)
)

# Normalize all to 2023/24 prices
trust_data <- trust_data |>
  mutate(
    unit_cost_paid = cost_paid_pounds / quantity,
    inflation_factor = mapply(
      function(from, to) nhscii(from_year = from, to_year = to, index = "prices"),
      from = purchase_year,
      to = "2023/24"
    ),
    normalized_unit_cost_2023_24 = unit_cost_paid * inflation_factor
  )

trust_data |>
  select(trust, purchase_year, unit_cost_paid, normalized_unit_cost_2023_24)
```

**Insight:** Unit cost increased from £0.15 to £0.22, but after normalizing all to 2023/24 prices, we see Trust A paid slightly *more* (in 2023/24 equivalent) — suggesting price changes or negotiation differences.

## Best practices

### 1. Always document your price year

```{r}
# Include metadata in your results
analysis_meta <- list(
  price_source = "NHS dm+d bundled data",
  price_year = "2023/24",
  inflation_index = "pay_and_prices",
  dmd_release = attr(dmd_master, "dmd_release_label"),
  analysis_date = Sys.Date()
)

str(analysis_meta)
```

### 2. Handle lookup failures gracefully

```{r}
# Function to safely look up prices with fallback
safe_lookup <- function(medicine_name, method = "fuzzy") {
  result <- dmd_price_lookup(
    medicine_name,
    method = method,
    max_dist = 3
  )
  
  if (nrow(result) == 0) {
    warning("No match found for: ", medicine_name)
    return(list(price = NA_real_, matched = FALSE))
  }
  
  list(
    price = result$basic_price[[1]],
    matched = TRUE,
    matched_name = result$medicine[[1]]
  )
}

safe_lookup("paracetamol 500mg tablets")
safe_lookup("xyz_medicine_not_real")
```

### 3. Separate data from analysis

```{r}
# Keep input data separate from calculated fields
analysis <- current_formulary |>
  select(medicine, current_monthly_packs, price_pence) |>
  mutate(
    # Derived calculations
    monthly_cost_pounds = (current_monthly_packs * price_pence) / 100,
    annual_cost_2023_24 = monthly_cost_pounds * 12,
    # Once 2024/25 rates are available, you can use:
    # inflated_2024_25 = annual_cost_2023_24 * nhscii("2023/24", "2024/25")
    .after = price_pence
  )
```

## Further reading

- [Costing principles guide (NHS England)](https://www.england.nhs.uk/publication/nhs-costing-standards/)
- [NICE cost-effectiveness assessment](https://www.nice.org.uk/process/pmg6)
- NHS CII vignette: `vignette("nhscii")`
