---
title: "Data Sources and Updates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Sources and Updates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`dmdprices` combines data from three authoritative NHS sources. This vignette explains where the data comes from, how often it updates, and how to stay current.

```{r setup}
library(dmdprices)
```

## The three data sources

### 1. NHS dm+d (Dictionary of Medicines and Devices)

**What it is:** The authoritative UK medicine and device dictionary, maintained by the NHSBSA.

**Update frequency:** Weekly (every Thursday)

**Content:**
- Virtual Medicinal Products (VMPs) — generic medicines
- Virtual Medicinal Product Packs (VMPPs) — packs/strengths
- Actual Medicinal Products (AMPs) — branded products
- SNOMED CT codes for each
- Links to Drug Tariff and NHS Indicative Prices

**Coverage:** All medicines approved for use in UK NHS

**Access:** NHSBSA TRUD (Technology Reference Data Update) service  
📍 https://isd.digital.nhs.uk/trud

**In dmdprices:**
```{r}
# The bundled dataset
data(dmd_master)

attr(dmd_master, "dmd_release_label")

nrow(dmd_master)
```

### 2. Drug Tariff

**What it is:** Official reimbursement prices for community pharmacies in England.

**Update frequency:** Monthly (published mid-month)

**Content:**
- Basic prices (pence)
- Drug Tariff part and category
- Uplift and discount schedules

**Published by:** NHSBSA

**Access:** NHSBSA website  
📍 https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff

**Important notes:**
- England only (Scotland/Wales have separate arrangements)
- Prices typically lower due to cost-containment policies
- Updated monthly; old prices may not reflect current reimbursement

**In dmdprices:** Column `basic_price`

### 3. NHS Indicative Prices

**What it is:** Reference prices for hospital and NHS secondary care purchasing.

**Update frequency:** Quarterly

**Content:**
- Indicative prices (pence)
- Basis of price (DT = Drug Tariff, MIMS, other)
- Date price became effective

**Published by:** NHS England

**Access:** NHS England Cost Collection  
📍 https://www.england.nhs.uk/publication/national-cost-collection-ncc/

**Important notes:**
- Indicative only (hospitals negotiate contracts)
- Broader supply chain reflected
- Usually higher than Drug Tariff (due to distribution markup)

**In dmdprices:** Columns `nhs_indicative_price`, `price_basis`, `price_date`

## NHS Cost Inflation Index (NHS CII)

**What it is:** Annual inflation rates for health and social care costs, published by the PSSRU.

**Update frequency:** Annually (usually in autumn/winter for the next year)

**Content:**
- Pay inflation rates
- Price inflation rates
- Combined pay+prices indices
- Coverage: Currently 2015/16–2023/24

**Published by:** Personal Social Services Research Unit (PSSRU), University of Kent

**Access:** PSSRU website  
📍 https://www.pssru.ac.uk/project-pages/unit-costs/

**In dmdprices:** Function `nhscii()`

```{r}
# Example: current rates available
nhscii("2015/16", "2023/24", index = "prices")
```

**Note:** 2023/24 figures are provisional. The 2025 PSSRU manual will update with additional data.

## How often should you update?

### Use bundled data if:
- You're doing academic analysis (reproducibility matters)
- Your project isn't time-sensitive
- You're learning the package

### Load fresh dm+d if:
- You're running operational systems (pharmacy, procurement)
- You need current Drug Tariff prices
- New medicines are important to your analysis
- More than 3 months have passed

### Update NHS CII if:
- You're inflating costs to future years
- New PSSRU manual has been published
- Your analysis year has changed

## Checking which versions you're using

### dm+d release

```{r}
# Check bundled data release
attr(dmd_master, "dmd_release_label")

# Check when it was packaged
attr(dmd_master, "package_date")
```

### When did you last download fresh data?

```{r}
# If you've loaded a fresh dm+d via dmd_load()
# Metadata is attached to the returned object
# db <- dmd_load("path/to/dmdDataLoader")
# attr(db, "load_date")
```

### NHS CII coverage

```{r}
# The rates currently available
# (from the .nhscii_rates internal object)
# Coverage: 2015/16 to 2023/24
# Use nhscii() with any year in that range

nhscii("2015/16", "2015/16")  # Should return 1
```

## Reproducibility and audit trails

### Document your data in every analysis

```{r}
# Create a data audit section in your report
data_audit <- list(
  analysis_date = Sys.Date(),
  
  dmd_source = list(
    source = "dmdprices bundled dataset",
    release = attr(dmd_master, "dmd_release_label"),
    medicines_total = nrow(dmd_master)
  ),
  
  inflation_source = list(
    source = "PSSRU Unit Costs of Health and Social Care",
    doi = "10.22024/UniKent/01.02.109563",
    coverage = "2015/16 to 2023/24 (2023/24 provisional)"
  )
)

str(data_audit)
```

### Save metadata with your results

```{r}
# Include in your output file
metadata <- data.frame(
  item = c(
    "Analysis date",
    "dm+d release",
    "NHS CII index",
    "Package version"
  ),
  value = c(
    as.character(Sys.Date()),
    attr(dmd_master, "dmd_release_label"),
    "pay_and_prices",
    as.character(packageVersion("dmdprices"))
  )
)

metadata
```

## Update workflow example

If you're doing regular analyses, here's a recommended workflow:

```r
# 1. Once per quarter: refresh dm+d
my_dm_d <- dmd_load("~/dmdDataLoader")  # Download fresh release
saveRDS(my_dm_d, "data/dm_d_current.rds")

# 2. Check if NHS CII needs updating
# Visit https://www.pssru.ac.uk and check if new manual is available
# (Current package covers through 2023/24; update when 2024/25 rates published)

# 3. In your analysis scripts:
dm_d <- readRDS("data/dm_d_current.rds")
medicine_cost <- dmd_price_lookup("Metformin 500mg", db = dm_d)

# 4. Record versions
analysis_metadata <- list(
  dm_d_release = attr(dm_d, "dmd_release_label"),
  cii_latest_year = "2023/24",  # Update when new rates available
  analysis_date = Sys.Date()
)
```

## Official data feeds (for developers)

If you're building on `dmdprices`, here are direct feeds:

### dm+d download
```
https://isd.digital.nhs.uk/trud/users/guest/filters/0/categories/6
```
(Free registration required)

### Drug Tariff CSV
```
https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff
```
(Monthly archive available)

### NHS Indicative Prices
```
https://www.england.nhs.uk/publication/national-cost-collection-ncc/
```

### PSSRU Unit Costs
```
https://www.pssru.ac.uk/project-pages/unit-costs/
```

## Licensing and attribution

### dm+d
© Crown copyright. Licensed under the Open Government Licence v3.0  
📍 https://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/

**Citation:**
```
NHS Business Services Authority (2025). Dictionary of Medicines and 
Devices (dm+d). Available: https://isd.digital.nhs.uk/trud
```

### Drug Tariff and NHS Indicative Prices
Published by NHSBSA and NHS England respectively. No special licence required for analysis (educational/research use).

**Citation:**
```
NHS Business Services Authority (2025). Drug Tariff. Available: 
https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff
```

### NHS CII
Published by PSSRU under Open Government Licence.

**Citation:**
```
PSSRU (2024). Unit Costs of Health and Social Care 2024. 
University of Kent. https://doi.org/10.22024/UniKent/01.02.109563
```

## Troubleshooting data issues

### "Why is this medicine missing?"

1. Check if it's in dm+d (might be new, delisted, or historical)
2. Try fuzzy matching with `method = "fuzzy"`
3. Search TRUD directly for the SNOMED code

### "Why is there no price?"

See vignette "Working with Drug Tariff and NHS Indicative Prices" for common reasons:
- Centrally procured (vaccines, infusions)
- Specialty/rare disease medicine
- Recently added to formulary
- Price confidentiality (some agreements)

### "Why does my price differ from the Drug Tariff?"

1. Different month/edition (Tariff updated monthly)
2. Bundled data may be 1-3 months old
3. Load fresh dm+d with `dmd_load()`

## Further reading

- NHS CII and inflation: `vignette("nhscii")`
- Working with prices: `vignette("drug_tariff_matching")`
- Cost analysis workflows: `vignette("cost_analysis_workflows")`
