---
title: "Getting started with dmdprices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dmdprices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dmdprices)
```

## Overview

`dmdprices` provides price lookups against the NHS Dictionary of Medicines and
Devices (dm+d). A bundled dataset (Week 34 2025) is included so no setup is
needed for immediate use.

There are two main functions:

| Function | Purpose |
|---|---|
| `dmd_price_lookup()` | Search for medicines by name and return pricing |
| `dmd_load()` | Load a more recent dm+d release from a local directory |

---

## Using the bundled data

### Partial match (default)

Returns all rows where the medicine name contains the query string
(case-insensitive):

```{r partial}
dmd_price_lookup("metformin")
```

### Exact match

Case-insensitive match against the full VMP name:

```{r exact}
dmd_price_lookup("Metformin 500mg tablets", method = "exact")
```

### Fuzzy match

Tolerates typos using optimal string alignment distance. Tune sensitivity with
`max_dist` (default `3`):

```{r fuzzy}
dmd_price_lookup("metfromin 500mg tablets", method = "fuzzy", max_dist = 4)
```

### Including products without prices

By default, rows where both `Basic Price` and `NHS Indicative Price` are `NA`
are dropped. Set `active_only = FALSE` to keep them:

```{r active-only}
dmd_price_lookup("metformin", active_only = FALSE)
```

### Checking the bundled release

```{r release}
attr(dmd_master, "dmd_release_label")
```

---

## Using a more recent release

If you have the NHSBSA dm+d extract tool (`dmdDataLoader`), load a newer
release with `dmd_load()` and pass the result to `dmd_price_lookup()`:

```{r load, eval = FALSE}
db <- dmd_load("path/to/dmdDataLoader")
dmd_price_lookup("metformin", db = db)
```

You can set a project-wide default path in your `.Rprofile` to avoid repeating
it each session:

```{r option, eval = FALSE}
options(dmdprices.path = "~/dmdDataLoader")
# Then simply:
db <- dmd_load()
```

---

## Output columns

Prices are in **pence**, matching the NHS Drug Tariff Part VIIIA CSV
convention. One row is returned per branded pack (AMPP).

| Column | Description |
|---|---|
| `Medicine` | VMP (generic) name |
| `Pack size` | Pack quantity |
| `Unit` | Unit of measure (tablet, ml, etc.) |
| `VMP Snomed Code` | VMP SNOMED CT identifier |
| `VMPP Snomed Code` | VMPP SNOMED CT identifier |
| `Drug Tariff Category` | e.g. "Part VIIIA Category M" |
| `Basic Price` | Drug Tariff basic price (pence) |
| `NHS Indicative Price` | NHS Indicative Price (pence) |
| `Price Basis` | Basis of NHS Indicative Price |
| `Price Date` | Date of NHS Indicative Price |
| `AMPP Name` | Branded pack name |
| `AMPP Snomed Code` | AMPP SNOMED CT identifier |

---

## Data attribution

The bundled `dmd_master` dataset is derived from the **NHS Dictionary of
Medicines and Devices (dm+d)**, Week 34 2025 (14 August 2025), published by
the **NHS Business Services Authority (NHSBSA)**.

© Crown copyright. Licensed under the
[Open Government Licence v3.0](https://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/).
