---
title: "Dose optimisation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dose optimisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dmdprices)
```

## Overview

`dmd_dose_optimise()` takes a clinical dose (e.g. 900 mg) and returns the
best combination of AMPPs (branded packs) from the dm+d that deliver it.
"Best" is defined by two objectives, returned side-by-side:

- **cheapest** — lowest pro-rata cost across priced combinations.
- **min_items** — fewest tablets, capsules, ampoules, or other discrete items.

Preparations are segregated automatically: immediate-release tablets,
modified-release tablets, oral solutions, and solutions for injection each
get their own row so they are never mixed in a single combination.

## A basic example

A 1,500 mg dose of metformin, restricted to standard (immediate-release)
tablets:

```{r, eval = FALSE}
res <- dmd_dose_optimise(
  "metformin",
  dose = 1500,
  dose_unit = "mg",
  preparation = "tablet|none|oral"
)
res
```

Each row contains a `combination` list-column identifying the specific
branded products picked:

```{r, eval = FALSE}
res$combination[[1]]
```

The combination tibble has one row per AMPP used, with the VMP name, branded
pack name, SNOMED codes, and the count / packs-to-buy / pro-rata /
whole-pack cost breakdown.

## When objectives differ

A 900 mg metformin dose is a good illustration. The cheapest combination
may use more items (several smaller tablets) while the minimum-items answer
may over-deliver (e.g. two 500 mg tablets → 1,000 mg). The output flags
`over_delivery` in both the column and the notes:

```{r, eval = FALSE}
res <- dmd_dose_optimise("metformin", dose = 900, dose_unit = "mg",
                         preparation = "tablet|none|oral")
res[, c("objective", "total_items", "dose_delivered",
        "over_delivery", "cost_prorata_pence", "cost_whole_pack_pence")]
```

## Segregating preparations

A morphine dose should never mix oral solutions with tablets or injections,
so `dmd_dose_optimise()` returns one result per preparation group:

```{r, eval = FALSE}
dmd_dose_optimise("morphine", dose = 20, dose_unit = "mg")[,
  c("preparation_label", "objective", "total_items",
    "dose_delivered", "over_delivery")]
```

## Micrograms

Small doses in micrograms are handled transparently. The requested dose
unit is retained in the output while the canonical form is used internally:

```{r, eval = FALSE}
dmd_dose_optimise("levothyroxine", dose = 125, dose_unit = "microgram")
```

## Pricing

By default, pro-rata costs (`pack_price × items ÷ pack_size`) are computed
from `basic_price`, with fallback to `nhs_indicative_price` per-AMPP when
the chosen column is NA (flagged in `notes`). Whole-pack costs
(`pack_price × packs_to_buy`) reflect what you would actually pay when you
cannot split a pack.

Switch the primary price field with `price = "nhs_indicative_price"`.

## Community pharmacy: whole-pack dispensing

In hospital dispensing, individual tablets can be taken from a part-pack
(`can_split = TRUE`, the default). In community pharmacy, whole packs must
be dispensed. Set `can_split = FALSE` to switch to pack-level optimisation:

```{r, eval = FALSE}
dmd_dose_optimise("metformin", dose = "1500 mg", can_split = FALSE)
```

When `can_split = FALSE`:

- The DP operates on whole packs rather than individual tablets.
- `dose_cost_pence` reflects the whole-pack cost (what the pharmacy
  actually pays), not a pro-rata fraction.
- The `notes` column includes `"no-pack-splitting"`.
- In the `combination` tibble, `count` is the number of **packs** dispensed
  rather than individual tablets.

Concentration-based preparations (liquids, vials, inhalers) are always
whole-container regardless of `can_split`, since one container is the
minimum dispensing unit.

## Vectorised costing with `dmd_dose_cost()`

For costing a column of doses inside a data frame, `dmd_dose_cost()` is
more efficient than calling `dmd_dose_optimise()` inside `purrr::map_dbl()`.
It accepts a numeric vector of doses and returns a numeric vector of costs in
pence of the same length:

```{r, eval = FALSE}
library(dplyr)

treatment_df |>
  mutate(
    ritux_cost_gbp = dmd_dose_cost(
      query       = "rituximab",
      dose        = dose_mg,
      dose_unit   = "mg",
      objective   = "cheapest",
      preparation = "infusion"
    ) / 100
  )
```

Key differences from `dmd_dose_optimise()`:

- `dose` must be a **numeric vector** (string doses are not accepted).
- Returns a plain `numeric` vector, not a tibble — no `combination` detail.
- The expensive candidate-preparation step runs **once** for all doses in
  the vector (memoized across calls within the same session).
- When multiple preparation groups match, the **minimum cost** across groups
  is returned. Specify `preparation` to restrict to a single route.
- `NA`, zero, or negative dose elements return `NA` (or a custom `na_value`).

## Cross-referencing with other dm+d data

The `combination` list-column includes `vmpp_snomed_code` and
`ampp_snomed_code` for each picked product, so you can join back to other
dm+d views or NHS Drug Tariff Part VIIIA CSVs.

## Limitations

- Discrete items only — tablet splitting is not considered.
- For liquids, each "item" is one container (bottle, ampoule, vial). A
  small dose requested against a large container is delivered as one whole
  container with an `over_delivery` flag.
- Compound products (multiple active ingredients in one VMP) are not
  optimised and are flagged in `notes`.
- Clinical safety (max single / max daily dose) is **not** enforced.
