---
title: "Dose optimisation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dose optimisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dmdprices)
```

## Overview

`dmd_dose_optimise()` takes a clinical dose (e.g. 1000 mg) and returns the
best combination of AMPPs (branded packs) from the dm+d that deliver it.
"Best" is controlled by the `objective` argument, which accepts one or more of:

- **cheapest** — lowest pro-rata cost across priced combinations.
- **min_items** — fewest tablets, capsules, ampoules, or other discrete items.
- **most_expensive** — highest-cost combination (useful for worst-case cost
  modelling or budget-impact analysis upper bounds).

Pass a character vector to request multiple objectives at once, or use
`objective = "all"` as a shorthand for all three. The default is
`c("cheapest", "min_items")`.

Preparations are segregated automatically: immediate-release tablets,
modified-release tablets, oral solutions, and solutions for injection each
get their own row so they are never mixed in a single combination.
Combination products (multiple active ingredients in one VMP) are skipped
with a warning because the dose target would be ambiguous — see
"Combination products" below for how to dose them by a single ingredient.
Packs holding several products (titration packs, co-packs) are skipped for
the same reason, with a warning of their own, when their name gives each
product's strength in mass or units.

## A basic example

A 1,500 mg dose of metformin, restricted to standard (immediate-release)
tablets (default objectives: cheapest + min_items):

```{r, eval = FALSE}
res <- dmd_dose_optimise(
  "metformin",
  dose = 1500,
  dose_unit = "mg",
  preparation = "tablet|none|oral"
)
res
```

Each row contains a `combination` list-column identifying the specific
branded products picked:

```{r, eval = FALSE}
res$combination[[1]]
```

The combination tibble has one row per AMPP used, with the VMP name, branded
pack name, SNOMED codes, and the count / packs-to-buy / pro-rata /
whole-pack cost breakdown.

## When objectives differ

A 1500 mg metformin dose is a good illustration. The cheapest combination may
use more items (several smaller tablets) while the minimum-items answer uses
fewer, larger ones. Both deliver 1500 mg exactly:

```{r, eval = FALSE}
res <- dmd_dose_optimise("metformin", dose = 1500, dose_unit = "mg",
                         preparation = "tablet|none|oral")
res[, c("objective", "total_items", "dose_delivered",
        "over_delivery", "cost_prorata_pence", "cost_whole_pack_pence")]
```

## Over-delivery

By default the optimiser returns only combinations that deliver the requested
dose **exactly** (`over_delivery = "forbid"`), because for tablets and capsules
a surplus is extra drug the patient takes. A preparation group that cannot hit
the dose exactly returns no row and is named in a warning, so an impossible
dose is never silently costed as an over-delivered one:

```{r, eval = FALSE}
# 750 mg cannot be built from 500 mg and 1,000 mg tablets — no rows, with a warning
dmd_dose_optimise("metformin", dose = 750, dose_unit = "mg",
                  preparation = "tablet|none|oral")

# smallest achievable over-delivery, objective applied within it
dmd_dose_optimise("metformin", dose = 750, dose_unit = "mg",
                  preparation = "tablet|none|oral",
                  over_delivery = "minimise")[,
  c("objective", "dose_delivered", "over_delivery", "dose_exact", "notes")]

# "allow" lets the objective decide across everything covering the dose
dmd_dose_optimise("metformin", dose = 750, dose_unit = "mg",
                  preparation = "tablet|none|oral",
                  over_delivery = "allow")
```

`dose_exact` flags whether the row delivers the dose exactly, so you do not have
to compare a floating-point `over_delivery` to zero.

The requested dose is never rounded to the strengths. A dose that no combination
of a group's strengths sums to — 2.4 mg against 1 mg tablets — has no exact
combination: the group returns no row under `"forbid"`, and under `"minimise"`
or `"allow"` the smallest combination delivering at least the dose (3 mg). No
result delivers less than the dose requested.

Whenever a returned combination over-delivers, one warning per call lists the
preparation groups affected and says which case it is — no exact combination
exists, or one exists but the objective preferred an over-delivering combination
(only possible under `"allow"`). These warnings are the only signal in
`dmd_dose_cost()` and `dmd_dose_cost_range()`, which return bare numbers with
no `notes` column to inspect. Pass `quiet = TRUE` to silence them in bulk
costing runs; unrelated warnings still come through:

```{r, eval = FALSE}
# warns: an exact 3 mg combination existed but the cheapest one over-delivers
dmd_dose_cost("buprenorphine", dose = 3, dose_unit = "mg",
              preparation = "sublingual tablet", over_delivery = "allow")

# same call, silenced
dmd_dose_cost("buprenorphine", dose = 3, dose_unit = "mg",
              preparation = "sublingual tablet", over_delivery = "allow",
              quiet = TRUE)
```

Two cases are **exempt**, because there the surplus stays in the pack or the
vial rather than reaching the patient, and the cheapest pack or container
covering the dose is the costing answer:

- whole-pack dispensing (`can_split = FALSE`), and
- whole containers — vials and ampoules with `can_split_vials = FALSE`.

Exempt rows carry an `"over-delivery-policy-not-applied"` note and do not warn,
since buying whole packs or vials to cover a dose is the expected costing
answer. Exact delivery from a container is available with
`can_split_vials = TRUE`.

The same argument is accepted by `dmd_dose_cost()` and `dmd_dose_cost_range()`,
where a dose that cannot be delivered exactly returns `NA` under the default.

## Segregating preparations

A morphine dose should never mix oral solutions with tablets or injections,
so `dmd_dose_optimise()` returns one result per preparation group:

```{r, eval = FALSE}
dmd_dose_optimise("morphine", dose = 20, dose_unit = "mg")[,
  c("preparation_label", "objective", "total_items",
    "dose_delivered", "over_delivery")]
```

## Micrograms

Small doses in micrograms are handled transparently. The requested dose
unit is retained in the output while the canonical form is used internally:

```{r, eval = FALSE}
dmd_dose_optimise("levothyroxine", dose = 125, dose_unit = "microgram")
```

## Pricing

By default, pro-rata costs (`pack_price × items ÷ pack_size`) are computed
from `basic_price`, with fallback to `nhs_indicative_price` per-AMPP when
the chosen column is NA (flagged in `notes`). Whole-pack costs
(`pack_price × packs_to_buy`) reflect what you would actually pay when you
cannot split a pack.

Switch the primary price field with `price = "nhs_indicative_price"`.

## Community pharmacy: whole-pack dispensing

In hospital dispensing, individual tablets can be taken from a part-pack
(`can_split = TRUE`, the default). In community pharmacy, whole packs must
be dispensed. Set `can_split = FALSE` to switch to pack-level optimisation:

```{r, eval = FALSE}
dmd_dose_optimise("metformin", dose = "1500 mg", can_split = FALSE)
```

When `can_split = FALSE`:

- The DP operates on whole packs rather than individual tablets.
- `dose_cost_pence` reflects the whole-pack cost (what the pharmacy
  actually pays), not a pro-rata fraction.
- The `notes` column includes `"no-pack-splitting"`.
- In the `combination` tibble, `count` is the number of **packs** dispensed
  rather than individual tablets.

With `can_split = TRUE`, concentration-based preparations (liquids, vials,
inhalers) are still costed in whole containers unless `can_split_vials = TRUE`,
since one bottle, vial, or ampoule is the minimum administration unit; with
`can_split = FALSE` they are optimised over whole packs like every other
preparation, so a pack of several syringes or ampoules is dispensed as whole
packs and `total_items` counts packs. For liquids where the pack quantity is in the same
unit as the concentration denominator (for example `10 mg/5 ml` in a `100 ml`
bottle), the optimiser uses the total pack volume to calculate the active
quantity in the container. A pack that counts containers (for example
`10 pre-filled disposable injection` or `5 ampoule`) is priced per container
— a tenth of the pack price for one of ten syringes — and its whole-pack cost
buys as many packs as the containers need.

## Worst-case cost

`objective = "most_expensive"` selects the highest-cost combination of AMPPs
that delivers the dose. This is useful for budget-impact analysis or scenario
modelling where you want an upper bound on expenditure:

```{r, eval = FALSE}
dmd_dose_optimise(
  "metformin",
  dose = 1000,
  dose_unit = "mg",
  preparation = "tablet|none|oral",
  objective = "most_expensive"
)
```

Request multiple objectives at once by passing a vector, or use `"all"` as
a shorthand for all three:

```{r, eval = FALSE}
# Three rows: cheapest, min_items, and most_expensive
dmd_dose_optimise(
  "metformin",
  dose = 1000,
  dose_unit = "mg",
  preparation = "tablet|none|oral",
  objective = "all"
)
```

## Vial sharing

By default, concentration-based preparations (vials, ampoules) are costed as
whole containers — even when the dose is a fraction of the container volume.
Set `can_split_vials = TRUE` to enable **vial sharing**, where the cost is
pro-rated to the exact fraction of the container needed:

```{r, eval = FALSE}
# 250 mg from a 500 mg/50 ml vial — costs half a vial, not a whole one
dmd_dose_optimise(
  "rituximab",
  dose = 250,
  dose_unit = "mg",
  preparation = "infusion",
  can_split_vials = TRUE
)
```

When vial sharing is active:

- The `count` in the `combination` tibble is the non-integer fraction of the
  container used (e.g. `0.5` for half a vial).
- `"vial-sharing"` is added to the `notes` column.
- The reported cost is ≤ the whole-container cost for the same dose.

This mode is appropriate when a single vial is shared between multiple patients
on the same day, as is common in day-case oncology units.

## Vectorised costing with `dmd_dose_cost()`

For costing a column of doses inside a data frame, `dmd_dose_cost()` is
more efficient than calling `dmd_dose_optimise()` inside `purrr::map_dbl()`.
It accepts a numeric vector of doses and returns a numeric vector of costs in
pence of the same length:

```{r, eval = FALSE}
library(dplyr)

treatment_df |>
  mutate(
    ritux_cost_gbp = dmd_dose_cost(
      query       = "rituximab",
      dose        = dose_mg,
      dose_unit   = "mg",
      objective   = "cheapest",
      preparation = "infusion"
    ) / 100
  )
```

Key differences from `dmd_dose_optimise()`:

- `dose` must be a **numeric vector** (string doses are not accepted).
- Returns a plain `numeric` vector, not a tibble — no `combination` detail.
- The expensive candidate-preparation step runs **once** for all doses in
  the vector (memoized across calls within the same session).
- When multiple preparation groups match, the **minimum cost** across groups
  is returned. Specify `preparation` to restrict to a single route.
- `NA`, zero, or negative dose elements return `NA` (or a custom `na_value`).

Use `dmd_dose_cost_range()` when you need both lower and upper cost bounds:

```{r, eval = FALSE}
dmd_dose_cost_range(
  "rituximab",
  dose = c(375, 500, 700),
  dose_unit = "mg",
  preparation = "infusion"
)
```

## Cross-referencing with other dm+d data

The `combination` list-column includes `vmpp_snomed_code` and
`ampp_snomed_code` for each picked product, so you can join back to other
dm+d views or NHS Drug Tariff Part VIIIA CSVs.

## Combination products

Some products contain more than one active ingredient — co-codamol
(codeine + paracetamol) is the familiar case, but the class also includes
allergen mix solutions and multi-factor concentrates whose names show only a
single number. A single dose target is ambiguous for these, so the optimiser
skips them and says so:

```{r, eval = FALSE}
dmd_dose_optimise("co-codamol", dose = 30, dose_unit = "mg")
#> Warning: 110 unsupported compound products skipped during dose optimisation.
#> x E.g. "Co-codamol 12.8mg/500mg tablets", ... and 9 more.
#> i Pass `ingredient = "<name>"` to dose one active ingredient of a
#>   combination product (e.g. the codeine in co-codamol).
```

A product counts as a combination when the dm+d `is_combination` flag says so —
derived from the dm+d virtual product ingredient (VPI) data and authoritative
even when the name reads as a single strength — or when its name lists
multiple active strengths.

A strength in brackets that only restates the product's strength in another
unit dimension is the same dose, so it does not make a product a combination:
eptacog alfa "1mg (50,000unit)" is dosed by mass like any single-strength
product. A bracketed strength in the same dimension, or one that names another
substance, is a competing dose basis and the product stays skipped:
"Iohexol 755mg/ml (Iodine 350mg/ml)" and
"Mexiletine hydrochloride 200mg (Mexiletine 167mg)" are two examples.

Packs that hold several products — titration packs such as
"Danicopan 50mg tablets and Danicopan 100mg tablets", and co-packs of
different products — have no single per-item strength either. They are
skipped with a separate warning that names the kind of pack; cost each product
in the pack individually instead. Here the single-strength tablets are still
optimised:

```{r, eval = FALSE}
dmd_dose_optimise("danicopan", dose = 200, dose_unit = "mg")
#> Warning: 1 multi-product pack skipped during dose optimisation.
#> x Multi-strength pack: "Danicopan 50mg tablets and Danicopan 100mg tablets".
#> i A pack holding several products has no single per-item strength; cost its
#>   products individually.
```

Detection relies on the name giving each product's strength in mass or units.
A co-pack whose other product is stated as a percentage, such as
"Fluconazole 150mg capsule and Clotrimazole 2% cream", is not recognised and
is costed as its first product. A pack whose name gives no strength, such as
"Generic Otezla tablets treatment initiation pack", is dropped like any other
product without a parsed strength, with no warning naming it.

Code that silences these warnings by matching their text should match
`"multi-product pack"` as well as `"unsupported compound product"`.

To dose a combination product, name the ingredient you are dosing against.
Candidates are then restricted to products containing that ingredient, and the
dose is matched against the ingredient's own dm+d strength rather than
whatever number appears in the product name:

```{r, eval = FALSE}
# 30 mg of codeine from co-codamol products, priced per preparation group
dmd_dose_optimise(
  "co-codamol", dose = 30, dose_unit = "mg",
  ingredient = "codeine phosphate"
)[, c("preparation_label", "objective", "dose_delivered",
      "dose_exact", "dose_cost_pence")]
```

Ingredient matching is case-insensitive and word-boundary based (so
`"codeine"` does not also match `"dihydrocodeine"`); see
`?dmd_dose_optimise` for the details, including the warnings raised when a
term matches several distinct ingredients or a non-mass strength.

## Limitations

- Discrete items only — tablet splitting is not considered.
- For liquids, each "item" is one container (bottle, ampoule, vial) unless
  `can_split_vials = TRUE`, or one whole pack when `can_split = FALSE`. A small
  dose requested against a large container (or pack) is
  delivered as one whole container with an `over_delivery` flag; containers are
  exempt from the `over_delivery` policy for that reason.
- Sprays, granules and similar products whose strength is per dose or
  actuation but whose pack is measured in ml or g (for example nicotine
  1mg/dose mouth spray in a 13.2 ml bottle, or lidocaine 10mg/dose spray in a
  50 ml bottle) are skipped with a warning: the dm+d records no dose count for
  such packs and the package does not invent one. The same applies when the
  pack and the strength are in different physical units (a 500micrograms/g
  shampoo in a 125 ml bottle). Cost them as whole packs with
  `dmd_price_lookup()`.
- Combination products (multiple active ingredients in one VMP) are skipped
  with a warning unless an `ingredient` is named — see "Combination products"
  above. Packs holding several products are skipped with a warning of their
  own when their name gives each product's strength in mass or units; a
  co-pack with a percentage-strength cream is costed as its first product.
- Clinical safety (max single / max daily dose) is **not** enforced.
