---
title: "Dose optimisation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dose optimisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dmdprices)
```

## Overview

`dmd_dose_optimise()` takes a clinical dose (e.g. 900 mg) and returns the
best combination of AMPPs (branded packs) from the dm+d that deliver it.
"Best" is controlled by the `objective` argument, which accepts one or more of:

- **cheapest** — lowest pro-rata cost across priced combinations.
- **min_items** — fewest tablets, capsules, ampoules, or other discrete items.
- **most_expensive** — highest-cost combination (useful for worst-case cost
  modelling or budget-impact analysis upper bounds).

Pass a character vector to request multiple objectives at once, or use
`objective = "all"` as a shorthand for all three. The default is
`c("cheapest", "min_items")`.

Preparations are segregated automatically: immediate-release tablets,
modified-release tablets, oral solutions, and solutions for injection each
get their own row so they are never mixed in a single combination.
Compound products with multiple active strengths in one VMP name are skipped
with a warning because the dose target would be ambiguous.

## A basic example

A 1,500 mg dose of metformin, restricted to standard (immediate-release)
tablets (default objectives: cheapest + min_items):

```{r, eval = FALSE}
res <- dmd_dose_optimise(
  "metformin",
  dose = 1500,
  dose_unit = "mg",
  preparation = "tablet|none|oral"
)
res
```

Each row contains a `combination` list-column identifying the specific
branded products picked:

```{r, eval = FALSE}
res$combination[[1]]
```

The combination tibble has one row per AMPP used, with the VMP name, branded
pack name, SNOMED codes, and the count / packs-to-buy / pro-rata /
whole-pack cost breakdown.

## When objectives differ

A 900 mg metformin dose is a good illustration. The cheapest combination may
use more items (several smaller tablets) while the minimum-items answer uses
fewer, larger ones. Both deliver 900 mg exactly:

```{r, eval = FALSE}
res <- dmd_dose_optimise("metformin", dose = 900, dose_unit = "mg",
                         preparation = "tablet|none|oral")
res[, c("objective", "total_items", "dose_delivered",
        "over_delivery", "cost_prorata_pence", "cost_whole_pack_pence")]
```

## Over-delivery

By default the optimiser returns only combinations that deliver the requested
dose **exactly** (`over_delivery = "forbid"`), because for tablets and capsules
a surplus is extra drug the patient takes. A preparation group that cannot hit
the dose exactly returns no row and is named in a warning, so an impossible
dose is never silently costed as an over-delivered one:

```{r, eval = FALSE}
# 750 mg cannot be built from 500 mg and 1,000 mg tablets — no rows, with a warning
dmd_dose_optimise("metformin", dose = 750, dose_unit = "mg",
                  preparation = "tablet|none|oral")

# smallest achievable over-delivery, objective applied within it
dmd_dose_optimise("metformin", dose = 750, dose_unit = "mg",
                  preparation = "tablet|none|oral",
                  over_delivery = "minimise")[,
  c("objective", "dose_delivered", "over_delivery", "dose_exact", "notes")]

# "allow" lets the objective decide across everything covering the dose
dmd_dose_optimise("metformin", dose = 750, dose_unit = "mg",
                  preparation = "tablet|none|oral",
                  over_delivery = "allow")
```

`dose_exact` flags whether the row delivers the dose exactly, so you do not have
to compare a floating-point `over_delivery` to zero.

Whenever a returned combination over-delivers, one warning per call lists the
preparation groups affected and says which case it is — no exact combination
exists, or one exists but the objective preferred an over-delivering combination
(only possible under `"allow"`). That warning is the only signal in
`dmd_dose_cost()` and `dmd_dose_cost_range()`, which return bare numbers with no
`notes` column to inspect. Pass `quiet = TRUE` to silence it in bulk costing
runs; unrelated warnings still come through:

```{r, eval = FALSE}
# warns: an exact 3 mg combination existed but the cheapest one over-delivers
dmd_dose_cost("buprenorphine", dose = 3, dose_unit = "mg",
              preparation = "sublingual tablet", over_delivery = "allow")

# same call, silenced
dmd_dose_cost("buprenorphine", dose = 3, dose_unit = "mg",
              preparation = "sublingual tablet", over_delivery = "allow",
              quiet = TRUE)
```

Two cases are **exempt**, because there the surplus stays in the pack or the
vial rather than reaching the patient, and the cheapest pack or container
covering the dose is the costing answer:

- whole-pack dispensing (`can_split = FALSE`), and
- whole containers — vials and ampoules with `can_split_vials = FALSE`.

Exempt rows carry an `"over-delivery-policy-not-applied"` note and do not warn,
since buying whole packs or vials to cover a dose is the expected costing
answer. Exact delivery from a container is available with
`can_split_vials = TRUE`.

The same argument is accepted by `dmd_dose_cost()` and `dmd_dose_cost_range()`,
where a dose that cannot be delivered exactly returns `NA` under the default.

## Segregating preparations

A morphine dose should never mix oral solutions with tablets or injections,
so `dmd_dose_optimise()` returns one result per preparation group:

```{r, eval = FALSE}
dmd_dose_optimise("morphine", dose = 20, dose_unit = "mg")[,
  c("preparation_label", "objective", "total_items",
    "dose_delivered", "over_delivery")]
```

## Micrograms

Small doses in micrograms are handled transparently. The requested dose
unit is retained in the output while the canonical form is used internally:

```{r, eval = FALSE}
dmd_dose_optimise("levothyroxine", dose = 125, dose_unit = "microgram")
```

## Pricing

By default, pro-rata costs (`pack_price × items ÷ pack_size`) are computed
from `basic_price`, with fallback to `nhs_indicative_price` per-AMPP when
the chosen column is NA (flagged in `notes`). Whole-pack costs
(`pack_price × packs_to_buy`) reflect what you would actually pay when you
cannot split a pack.

Switch the primary price field with `price = "nhs_indicative_price"`.

## Community pharmacy: whole-pack dispensing

In hospital dispensing, individual tablets can be taken from a part-pack
(`can_split = TRUE`, the default). In community pharmacy, whole packs must
be dispensed. Set `can_split = FALSE` to switch to pack-level optimisation:

```{r, eval = FALSE}
dmd_dose_optimise("metformin", dose = "1500 mg", can_split = FALSE)
```

When `can_split = FALSE`:

- The DP operates on whole packs rather than individual tablets.
- `dose_cost_pence` reflects the whole-pack cost (what the pharmacy
  actually pays), not a pro-rata fraction.
- The `notes` column includes `"no-pack-splitting"`.
- In the `combination` tibble, `count` is the number of **packs** dispensed
  rather than individual tablets.

Concentration-based preparations (liquids, vials, inhalers) are treated as one
container regardless of `can_split`, since one bottle, vial, or ampoule is the
minimum dispensing unit. For liquids where the pack quantity is in the same
unit as the concentration denominator (for example `10 mg/5 ml` in a `100 ml`
bottle), the optimiser uses the total pack volume to calculate the active
quantity in the container.

## Worst-case cost

`objective = "most_expensive"` selects the highest-cost combination of AMPPs
that delivers the dose. This is useful for budget-impact analysis or scenario
modelling where you want an upper bound on expenditure:

```{r, eval = FALSE}
dmd_dose_optimise(
  "metformin",
  dose = 1000,
  dose_unit = "mg",
  preparation = "tablet|none|oral",
  objective = "most_expensive"
)
```

Request multiple objectives at once by passing a vector, or use `"all"` as
a shorthand for all three:

```{r, eval = FALSE}
# Three rows: cheapest, min_items, and most_expensive
dmd_dose_optimise(
  "metformin",
  dose = 1000,
  dose_unit = "mg",
  preparation = "tablet|none|oral",
  objective = "all"
)
```

## Vial sharing

By default, concentration-based preparations (vials, ampoules) are costed as
whole containers — even when the dose is a fraction of the container volume.
Set `can_split_vials = TRUE` to enable **vial sharing**, where the cost is
pro-rated to the exact fraction of the container needed:

```{r, eval = FALSE}
# 250 mg from a 500 mg/50 ml vial — costs half a vial, not a whole one
dmd_dose_optimise(
  "rituximab",
  dose = 250,
  dose_unit = "mg",
  preparation = "infusion",
  can_split_vials = TRUE
)
```

When vial sharing is active:

- The `count` in the `combination` tibble is the non-integer fraction of the
  container used (e.g. `0.5` for half a vial).
- `"vial-sharing"` is added to the `notes` column.
- The reported cost is ≤ the whole-container cost for the same dose.

This mode is appropriate when a single vial is shared between multiple patients
on the same day, as is common in day-case oncology units.

## Vectorised costing with `dmd_dose_cost()`

For costing a column of doses inside a data frame, `dmd_dose_cost()` is
more efficient than calling `dmd_dose_optimise()` inside `purrr::map_dbl()`.
It accepts a numeric vector of doses and returns a numeric vector of costs in
pence of the same length:

```{r, eval = FALSE}
library(dplyr)

treatment_df |>
  mutate(
    ritux_cost_gbp = dmd_dose_cost(
      query       = "rituximab",
      dose        = dose_mg,
      dose_unit   = "mg",
      objective   = "cheapest",
      preparation = "infusion"
    ) / 100
  )
```

Key differences from `dmd_dose_optimise()`:

- `dose` must be a **numeric vector** (string doses are not accepted).
- Returns a plain `numeric` vector, not a tibble — no `combination` detail.
- The expensive candidate-preparation step runs **once** for all doses in
  the vector (memoized across calls within the same session).
- When multiple preparation groups match, the **minimum cost** across groups
  is returned. Specify `preparation` to restrict to a single route.
- `NA`, zero, or negative dose elements return `NA` (or a custom `na_value`).

Use `dmd_dose_cost_range()` when you need both lower and upper cost bounds:

```{r, eval = FALSE}
dmd_dose_cost_range(
  "rituximab",
  dose = c(375, 500, 700),
  dose_unit = "mg",
  preparation = "infusion"
)
```

## Cross-referencing with other dm+d data

The `combination` list-column includes `vmpp_snomed_code` and
`ampp_snomed_code` for each picked product, so you can join back to other
dm+d views or NHS Drug Tariff Part VIIIA CSVs.

## Limitations

- Discrete items only — tablet splitting is not considered.
- For liquids, each "item" is one container (bottle, ampoule, vial) unless
  `can_split_vials = TRUE`. A small dose requested against a large container is
  delivered as one whole container with an `over_delivery` flag; containers are
  exempt from the `over_delivery` policy for that reason.
- Compound products (multiple active ingredients in one VMP) are skipped with
  a warning and are not returned as optimiser rows.
- Clinical safety (max single / max daily dose) is **not** enforced.
