---
title: "Working with Drug Tariff and NHS Indicative Prices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Drug Tariff and NHS Indicative Prices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The NHS maintains two key price lists for medicines:

- **Drug Tariff**: Official reimbursement prices for community pharmacies (updated monthly)
- **NHS Indicative Prices**: List prices for hospitals and other NHS settings (quarterly updates)

The `dmdprices` package unifies these sources with the **dm+d** (Dictionary of Medicines and Devices), allowing you to reconcile data from both lists seamlessly.

```{r setup}
library(dmdprices)
```

## Understanding the price columns

The bundled `dmd_master` dataset includes columns from both sources:

```{r}
# Load the bundled data
data(dmd_master)

# Examine a small subset
dmd_master |>
  dplyr::select(
    medicine, pack_size, unit,
    basic_price, nhs_indicative_price, price_basis
  ) |>
  dplyr::slice(1:3)
```

**Column meanings:**

| Column | Source | Meaning |
|--------|--------|---------|
| `basic_price` | Drug Tariff | Community pharmacy reimbursement (pence) |
| `nhs_indicative_price` | NHS Indicative | Hospital/NHS typical price (pence) |
| `price_basis` | NHS Indicative | Basis of the indicative price (e.g. "DT", "MIMS") |
| `price_date` | NHS Indicative | Date price was current |

## Key differences between the two lists

### Drug Tariff prices

- Updated **monthly** (typically mid-month)
- Legally binding for pharmacy reimbursement
- Usually lower (patient cost-containment)
- Available for most community medicines
- Column: `basic_price`

### NHS Indicative Prices

- Updated **quarterly**
- Reference only (hospitals negotiate)
- May be higher than Drug Tariff
- Reflects broader supply chain
- Column: `nhs_indicative_price`

### When prices differ

```{r}
# Find medicines where Drug Tariff < NHS Indicative
price_discrepancies <- dmd_master |>
  dplyr::filter(!is.na(basic_price), !is.na(nhs_indicative_price)) |>
  dplyr::mutate(
    tariff_cheaper = basic_price < nhs_indicative_price,
    pct_difference = ((nhs_indicative_price - basic_price) / basic_price) * 100
  ) |>
  dplyr::filter(tariff_cheaper) |>
  dplyr::arrange(dplyr::desc(pct_difference))

price_discrepancies |>
  dplyr::select(medicine, basic_price, nhs_indicative_price, pct_difference) |>
  dplyr::slice(1:5)
```

This is **normal**: the Drug Tariff is typically more aggressive on cost control for high-volume, low-cost items.

## Handling missing prices

Not all medicines have prices in both lists:

```{r}
# Medicines with Drug Tariff but no NHS Indicative
tariff_only <- dmd_master |>
  dplyr::filter(!is.na(basic_price), is.na(nhs_indicative_price))

nrow(tariff_only)

# Medicines with NHS Indicative but no Drug Tariff (often specialty drugs)
indicative_only <- dmd_master |>
  dplyr::filter(is.na(basic_price), !is.na(nhs_indicative_price))

nrow(indicative_only)
```

### Centrally-funded products

Some NHS medicines are centrally procured and have no retail price:

```{r}
# These typically appear as missing in both lists
no_price <- dmd_master |>
  dplyr::filter(is.na(basic_price), is.na(nhs_indicative_price))

nrow(no_price)
```

These include:
- Vaccines (centrally procured)
- Infusions and injectables (often bespoke agreements)
- Orphan/rare disease medicines
- Some biosimilars

**For analysis:** decide upfront whether to exclude, estimate, or flag these records.

## Matching medicines between sources

If you have a **Drug Tariff CSV** from NHS, use `dmd_price_lookup()` to reconcile:

```{r}
# Example: you have a list of medicine names from your pharmacy system
your_medicines <- c("Metformin 500mg tablets", "Lisinopril 10mg tablets")

# Look them up in dm+d
results <- lapply(your_medicines, dmd_price_lookup, method = "exact")
results
```

### Fuzzy matching for typos/variations

If exact matches fail, use fuzzy matching:

```{r}
# Your source data might have minor spelling variations
dmd_price_lookup("metformin 500 mg tablets", method = "fuzzy", max_dist = 3)
```

## Reconciling external price lists

If you're importing a Drug Tariff extract:

```{r}
# Simulated Drug Tariff data
external_tariff <- data.frame(
  medicine_name = c("Paracetamol 500mg tablets", "Ibuprofen 200mg tablets"),
  external_price_pence = c(150, 280)
)

# Match each medicine to dm+d
external_tariff$vmp_name <- sapply(
  external_tariff$medicine_name,
  function(x) {
    result <- dmd_price_lookup(x, method = "fuzzy", max_dist = 2)
    if (nrow(result) > 0) result$medicine[[1]] else NA_character_
  }
)

# Now you can compare prices
external_tariff
```

## Best practices

### 1. Always check for missing prices first

```{r}
# Flag records with incomplete pricing
dmd_master |>
  dplyr::mutate(
    has_tariff = !is.na(basic_price),
    has_indicative = !is.na(nhs_indicative_price),
    pricing_status = dplyr::case_when(
      has_tariff & has_indicative ~ "Both sources",
      has_tariff ~ "Tariff only",
      has_indicative ~ "Indicative only",
      TRUE ~ "No price"
    )
  ) |>
  dplyr::count(pricing_status)
```

### 2. Document your price source choice

```{r}
# Create a unified price column with priority logic
analysis_data <- dmd_master |>
  dplyr::mutate(
    analysis_price_pence = dplyr::coalesce(basic_price, nhs_indicative_price),
    price_source = dplyr::case_when(
      !is.na(basic_price) ~ "Drug Tariff",
      !is.na(nhs_indicative_price) ~ "NHS Indicative",
      TRUE ~ "Missing"
    )
  )

# Report this in your methodology
analysis_data |>
  dplyr::count(price_source)
```

### 3. Note the currency and units

Prices in `dmdprices` are in **pence**. Always convert for reporting:

```{r}
# Convert to pounds for reports
analysis_data |>
  dplyr::mutate(price_pounds = analysis_price_pence / 100) |>
  dplyr::select(medicine, price_pence = analysis_price_pence, price_pounds) |>
  dplyr::slice(1:3)
```

### 4. Keep audit trails

```{r}
# Record which version of dm+d was used
analysis_timestamp <- list(
  dmd_release = attr(dmd_master, "dmd_release_label"),
  analysis_date = Sys.Date()
)

analysis_timestamp
```

## Further reading

- [NHSBSA Drug Tariff](https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff)
- [NHS Indicative Prices](https://www.england.nhs.uk/publication/national-cost-collection-ncc/)
- [dm+d on TRUD](https://isd.digital.nhs.uk/trud)
