---
title: "NHS Cost Inflation Index (NHS CII)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NHS Cost Inflation Index (NHS CII)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The NHS Cost Inflation Index (NHS CII) is published annually by the PSSRU (Personal Social Services Research Unit) as part of their *Unit Costs of Health and Social Care* publication. It provides standardized rates to adjust health and social care costs for inflation across financial years.

This vignette shows how to use `nhscii()` and `inflate_nhscii()` to work with NHS CII rates in your analyses.

```{r setup}
library(dmdprices)
```

## Getting inflation factors

Use `nhscii()` to compute the inflation adjustment between two financial years:

```{r}
# Inflate from 2019/20 to 2023/24
nhscii("2019/20", "2023/24")
```

A factor of `1.127` means a 12.7% increase over that period.

### Using numeric years

For convenience, you can also pass numeric years (interpreted as end-years):

```{r}
# 2020 is interpreted as "2019/20", 2024 as "2023/24"
nhscii(2020, 2024)
```

### Percentage change

To get the percentage change directly:

```{r}
nhscii("2019/20", "2023/24", output_type = "percent")
```

### Different indices

Three indices are available: `"pay_and_prices"` (default), `"pay"`, and `"prices"`.

```{r}
# Pay component only
nhscii("2021/22", "2023/24", index = "pay")

# Prices component only
nhscii("2021/22", "2023/24", index = "prices")
```

Notice that pay inflation (`7.41%`) was higher than general price inflation (`7.15%`) in 2022/23.

## Adjusting costs

Use `inflate_nhscii()` to adjust cost values to a different financial year:

```{r}
# A single cost
inflate_nhscii(1000, "2019/20", "2023/24")
```

### Multiple costs at once

```{r}
costs <- c(150, 500, 2000)
inflated <- inflate_nhscii(
  costs,
  from_year = "2019/20",
  to_year = "2023/24",
  index = "pay_and_prices"
)

data.frame(
  original = costs,
  inflated_2023_24 = inflated
)
```

## Real-world example

Suppose you have unit costs for a health intervention from 2020/21 and need to report them in 2023/24 prices:

```{r}
# Unit costs in 2020/21
unit_cost_2020_21 <- 450

# Inflate to 2023/24
unit_cost_2023_24 <- inflate_nhscii(
  unit_cost_2020_21,
  from_year = 2021,
  to_year = 2024,
  index = "pay_and_prices"
)

unit_cost_2023_24
```

If you were comparing across multiple sectors (e.g., staff costs vs. service delivery), you might use different indices:

```{r}
staff_cost <- 200
non_staff_cost <- 250

# Staff costs inflate with pay index
staff_inflated <- inflate_nhscii(staff_cost, 2021, 2024, index = "pay")

# Non-staff costs inflate with prices index
non_staff_inflated <- inflate_nhscii(non_staff_cost, 2021, 2024, index = "prices")

data.frame(
  component = c("Staff", "Non-staff"),
  original = c(staff_cost, non_staff_cost),
  inflated = c(staff_inflated, non_staff_inflated)
)
```

## Data source and provenance

The NHS CII rates are derived from:

Jones KC, Weatherly H, Birch S, Castelli A, Chalkley M, Dargan A, Findlay D,
Gao M, Hinde S, Markham S, Smith D, Teo H (2025). *Unit Costs of Health and
Social Care 2024 Manual*. Technical report. Personal Social Services Research
Unit (University of Kent) & Centre for Health Economics (University of York),
Kent, UK. <https://doi.org/10.22024/UniKent/01.02.109563>

Licensed under [CC BY-NC-SA 4.0](https://creativecommons.org/licenses/by-nc-sa/4.0/).

Current coverage: 2015/16 to 2023/24

### A note on 2023/24 figures

The 2023/24 values in this package are **provisional**. As is standard for each annual PSSRU publication, later releases (e.g., the 2025 manual) may revise these figures when additional data become available. Check the latest PSSRU manual for the most current rates.

## Available years

To see which financial years are currently supported:

```{r}
# Pay and prices index
nhscii("2015/16", "2015/16")  # Check earliest available year

# All indices cover the same period:
# 2015/16, 2016/17, 2017/18, 2018/19, 2019/20, 2020/21, 2021/22, 2022/23, 2023/24
```

If you try to use an unavailable year, you'll get a helpful error:

```{r, error = TRUE}
nhscii("2014/15", "2023/24")
```

## Further reading

- [PSSRU Unit Costs of Health and Social Care](https://www.pssru.ac.uk/project-pages/unit-costs/)
- [NHS costing guidance](https://www.england.nhs.uk/publication/nhs-costing-standards/)
